The Units of Permissionless Consensus: \\ Towards Mobile and Edge Computing

### 2022-10-11 theoretical-consensus/article/article.tex
page 02

%\usepackage{algorithmic}

@@

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multi-agent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The establishment of a definition for the problem of reaching agreement in a distributed 
system was pioneered by Lamport et al. in \cite{pease1980reaching}. 
The authors defined consensus as the problem of agreeing on a single value among a set of processes, 
in the presence of faulty entities. The first consensus algorithms were designed for 
synchronous systems, where the communication between the processes is reliable, and the 
delay is bounded. However, these initial attempts failed to cover the different types of 
faulty behaviour. Along with the establishment of the famous Byzantine Generals Problem, 
the first solutions, not only for dealing with treason, but also for unreliable communication 
channels, or any other kind of arbitrary byzantine behaviour, were 
also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
mechanism that used a set of leaders to reach consensus. Multiple practical implementations 
and optimizations to this solution have been proposed in the literature.

@@

The advancements of the internet 
more than potentiated the revolution and what we now call the permissionless consensus problem was 
finally born. Without forgetting the previous attempts, the first practical permissionless consensus algorithm was 
proposed by Nakamoto in \cite{nakamoto2008bitcoin}. It is a proof-of-work consensus protocol that resembles a "replicated 
state machine" where the independent participants reach agreement not only about transactional values, 
but also about their order - naturality forming the underlying structure of what is now known as a blockchain. 
"Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
based on different consensus units, like proof-of-stake, proof-of-space, proof-of-burn, etc. The chaotic 
diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, 8629877, 9376868, BOURAGA2021114384}. 
The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8661654, 8168250, SALIMITARI2020100212}. 
This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
consensus protocols in mobile and edge devices.

@@

Following the line of the previous sections, it has been reasoned about the need 
for permissionless consensus when there are already well known and established consensus protocols
that work in trusted environments \cite{castro1999practical}. However, even those protocols have
their own limitations, not only in terms of trust, fault-tolerance, centrality, permissioning,
or bottlenecking, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
in size. The more participants there are, the more messages need to be exchanged
between them, and the more time it takes to reach consensus, even if assuring deterministic finality \cite{decker2016bitcoin}. 
This is a problem that is not only related to the number of participants, but also to the
communication fashion and bandwidth, and the computational capacity of the devices that
participate in the consensus protocol. 
Summing up, the need for permissionless consensus is then justified by the fact that
permissioned protocols are not compatible with the requirements of the new generation of
distributed systems, especially in the context of blockchain networks. 
These requirements include dealing with today's sparse networks of anonymously and dynamically
participating devices, without interrupting consensus and while generally battling Sybil attacks \cite{survey-dist-consensus}.
Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
can be run in a "zero-trust" environment, where the participants are unknown and untrusted,
and where the network is sparse and unpredictably unreliable.

@@

Technically, permissionless environments allow for larger networks
that depit lower connectivity between the participants. Operationally,
everything is expected to happen in an asynchronous or partially synchronous fashion, 
and the number of transactions is expected to be smaller than the 
permissioned counterparts. Nevertheless, participation is free, and the
governance is not centralized, but rather distributed and public. 
The identity of the participants is secured or semi-secure as it often relies on
pseudonymity for protecting the nodes' identity, while enabling full transparency 
in regards to the rest of the network content and operation \cite{xiao2019distributed}.
The goal of permissionless consensus is, as generally for any consensus protocol, to reach
agreement on a single value, or a set of values. However, due to the nature of the
protocols, the values that are agreed upon end up establishing the serialization of the
transactions that are executed in the network, and so establishing time consciousness and 
total order of the events.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by interchaining the generation of the agreeable value, 
in this particular case, a block and its proof, with the dissemination of 
the information to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and verifiable proof.
All of the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

>>>>

\usepackage{algorithm}
\usepackage{algpseudocode}

@@

\algdef{SE}[SUBALG]{Indent}{EndIndent}{}{\algorithmicend\ }%
\algtext*{Indent}
\algtext*{EndIndent}

\newcommand{\LineComment}[1]{\State \texttt{/* #1 */}}

@@

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{L}{ong} has been the time when consensus started 
to be defined as a fundamental problem of distributed systems \cite{pease1980reaching, lamport2019byzantine, lamport1983weak}. 
Generally, consensus means reaching an agreement between multiple parties 
in the potential presence of faulty individuals. As per multiagent systems, 
interacting over computer networks, consensus is thought to be the result 
of a coordination effort such that those parties agree on some value at a 
given moment. Achieving consensus implies that the system shall be reliable 
and fault-tolerant. However, the consensus problem has been limited by some 
assumptions on the networks. The well-known "secure Byzantine-Fault-Tolerant 
multiparty consensus systems" that have been designed over the years are usually 
meant to work only with a set of known participants, faulty or not \cite{castro1999practical}. The 
other side of the coin is the permissionless consensus challenge, consisting of 
achieving agreement in an environment where the participants are unknown and 
untrusted \cite{nakamoto2008bitcoin, buterin2014next}. Plus, there are other intrinsic particularities of 
this type of networks, for example, their openness, and the lack of any kind of central 
authority. This adds another layer of complexity to the problem, as the participants 
are not only unknown and untrusted but can also join or leave the network at any time, 
freely choosing if they want to participate in the consensus protocol or not. Nevertheless, 
the problem of permissionless consensus can still be seen as a special case of the more 
general consensus and can still be formalized in the same way. In this paper, we will 
focus on consensus in permissionless systems, especially in the context of blockchain 
networks. We will reason about its meaningfulness, ultimately by trying to identify the units that 
may underpin consensus. We will also discuss the current state-of-the-art, with a 
particular interest in the shift of the consensus layer of these distributed networks 
to mobile and edge computing environments, for which computationally expensive consensus 
algorithms are impractical and unfair.

@@

The establishment of a definition for the problem of reaching agreement in a distributed 
system was pioneered by Lamport et al. in \cite{pease1980reaching}. 
The authors defined consensus as the problem of agreeing on a single value among a set of processes, 
in the presence of faulty entities. The first consensus algorithms were designed for 
synchronous systems, where the communication between the processes is reliable, and the 
delay is bounded. However, these initial attempts failed to cover the different types of 
faulty behaviour. Along with the establishment of the famous Byzantine Generals Problem, 
the first solutions, not only for dealing with treason, but also for unreliable communication 
channels, or any other kind of arbitrary Byzantine behaviour, were 
also proposed by Lamport in \cite{lamport2019byzantine,lamport1983weak}. The solution was a synchronous 
mechanism that used a set of leaders to reach consensus. Multiple practical implementations 
and optimizations to this solution have been proposed in the literature.

@@

The advancements of the internet 
more than potentiated the revolution and what we now call the permissionless consensus problem was 
finally born. Without forgetting the previous attempts, the first practical permissionless consensus algorithm was 
proposed by Nakamoto in \cite{nakamoto2008bitcoin}. It is a proof-of-work consensus protocol that resembles a "replicated 
state machine" where the independent participants reach agreement not only about transactional values, 
but also about their order - naturally forming the underlying structure of what is now known as a blockchain. 
"Proof-of-work is essentially one-CPU-one-vote" and this is the novelty 
introduced by Bitcoin \cite{pass2016hybrid,pass2016hybrid2}. The focus shifted 
for decentralized systems and after proof-of-work many other consensus mechanisms have been proposed, 
based on different consensus units, like proof-of-stake, proof-of-space, proof-of-burn, etc. The chaotic 
diversity of new consensus protocols gave also room for endless reviews, overviews and comparisons \cite{survey-dist-consensus, BAMAKAN2020113385, 8400278, 8629877, 9376868, BOURAGA2021114384}. 
The authors of these surveys often put multiple dimensions into comparison, like fault tolerance, scalability, or energy consumption,
and, among those, some focused their efforts on mechanisms that may work in resource-constrained networks \cite{9104713, 8661654, 8168250, SALIMITARI2020100212}. 
This paper will try to identify the common conclusions from these comparisons, while looking into the state-of-the-art and novel approaches for running permissionless
consensus protocols in mobile and edge devices.

@@

Following the line of the previous sections, it has been reasoned about the need 
for permissionless consensus when there are already well known and established consensus protocols
that work in trusted environments \cite{castro1999practical, miller2016honey}. However, even those protocols have
their own limitations, not only in terms of trust, fault-tolerance, centrality, permissions,
or bottlenecks, but also in terms of scalability \cite{miller2016honey}. Trying to put some effort on decentralization, 
Byzantine-Fault-Tolerant consensus protocols are not known for their performance when the network grows
in size. The more participants there are, the more messages need to be exchanged
between them, and the more time it takes to reach consensus, even if assuring deterministic finality \cite{decker2016bitcoin}. 
This is a problem that is not only related to the number of participants, but also to the
communication fashion and bandwidth, and the computational capacity of the devices that
participate in the consensus protocol. 
Summing up, the need for permissionless consensus is then justified by the fact that
permissioned protocols are not compatible with the requirements of the new generation of
distributed systems, especially in the context of blockchain networks. 
These requirements include dealing with today's sparse networks of anonymously and dynamically
participating devices, without interrupting consensus and while battling Sybil attacks \cite{survey-dist-consensus}.
Fundamentally, the permissionless consensus problem is the need for a consensus protocol that
can be run in a distributed and decentralized environment, where the participants are unknown and untrusted,
and where the network is sparse and unpredictably unreliable.

@@

Technically, permissionless environments allow for larger networks
that depict lower connectivity between the participants. Operationally,
everything is expected to happen in an asynchronous or partially synchronous fashion, 
and the number of transactions is expected to be smaller than the 
permissioned counterparts. Nevertheless, participation is free, and the
governance is not centralized, but rather distributed and public. 
The identity of the participants is secured or semi-secure as it often relies on
pseudonymity for protecting the nodes' identity, while enabling full transparency 
in regards to the rest of the network content and operation \cite{xiao2019distributed}.
The goal of permissionless consensus, as generally for any consensus protocol, is to reach
agreement on a single value, or a set of values. However, due to the nature of the
protocols, the values that are agreed upon end up establishing the serialization of the
transactions that are executed in the network, and so establishing time consciousness and 
total order of the events.

@@

Also described in \cite{survey-dist-consensus}, very concisely, the 
way to achieve an operating protocol, as seen in the mainstream blockchain
networks, is by chaining the generation of the agreeable value, 
in this particular case, a block and its proof, with the dissemination of 
the information to the rest of the network, followed by the eventual validation and 
acceptance of the block by the rest of the nodes. This is the moment when consensus is reached.
Nevertheless, during the whole process, a fair and somewhat predictable incentive mechanism 
is also needed, that rewards the participants for their honest effort in reaching consensus, and
punishes the ones that are not behaving correctly. These incentives are of major importance
in this very context of permissionless consensus. 
These building blocks form the basis of the inner functioning of Bitcoin itself \cite{nakamoto2008bitcoin},
and are replicated with some variations in the other permissionless networks \cite{buterin2014next}.
For example, there are some networks that do not have a proof-of-work mechanism,
but rather a proof-of-stake, or a proof-of-space, or a proof-of-burn mechanism, etc.
when it comes to the generation of the block and its existential and later verifiable proof.
All of the other pillars are generally the same \cite{survey-dist-consensus}.
The next sections will try to give a more detailed overview of these 
block generation proof units.

\section{Proof-of-Work as a reference}

\subsection{The block generation}

In the classical Nakamoto consensus protocol, the generation of a block, to be proposed
for further network agreement, complies with the unit of computational work needed to
create, or rather find, a verifiable proof of the effort spent on creating the block \cite{nakamoto2008bitcoin}. 
This essentially requires brute forcing the finding of a cryptographic hash value for the
aggregation of the block information with a nonce such that this hash value satisfies
a difficulty threshold, which gets adjusted dynamically over time, to maintain the network overall 
requirement for the block generation interval \cite{survey-dist-consensus}.

Mitigation of Sybil attacks

\begin{algorithm}
\caption[short]{BlockGeneration}
\begin{algorithmic}[1]
  \Function {}{}
    \State $Block Header \gets Transaction \ Merkle \ Tree \ Root$
    \Indent
      \State $| \ Hash \ of \ the \ last \ Block$
      \State $| \ Timestamp$
      \State $| \ Other$
    \EndIndent

    \LineComment{the preceding zero bits in $target$ depict the mining difficulty}
    \While{$Hash(BlockHeader \ | \ nonce) \geq target$}
    \State Increment $nonce$
    \EndWhile
    \State \Return new block
  \EndFunction
\end{algorithmic}
\end{algorithm}
